---
title: 'COVID-19 Severity Phenotype: characteristics of individuals'
author: 'Chris Tomlinson'
date: 'Knitted on `r date()`'

output:
  html_document:
    df_print: paged
---

**Description** 

This notebook aggregates the data within `ccu013_paper_table_one_56million_denominator` and stratifies by `severity` which represents a picture of the most severe healthcare presentation per patient (mutually-exclusive).
  
**Project(s)** CCU013

**Paper** Characterising COVID-19 related events in a nationwide electronic health record cohort of 55.9 million people in England
 
**Author(s)** Chris Tomlinson
 
**Reviewer(s)** 
 
**Date last updated** 2021-10-21
 
**Date last reviewed** *NA*
 
**Date last run** `r date()`
 
**Data input**  
* `ccu013_paper_table_one_56million_denominator`

**Data output**  
* Self-contained as the knitted `.html` file from this notebook

**Software and versions** `SQL`, `R`
 
**Packages and versions**

```{r config}
# Load libraries prior to sessionInfo() to ensure documented
library(dplyr)
library(tidyr)
library(magrittr)
library(stringr)
library(tableone)

warning("Requires authenticated connection to Spark saved as 'con'")
source("/mnt/efs/c.tomlinson/dbConnect.R")
```

```{r sessionInfo}
sessionInfo()
```


## 1. Load & process cohort

```{r load}
# First load COVID patients
cohort_covid = dbGetQuery(con, 
                    "SELECT
                      death_covid,
                      death_all,
                      severity,
                      sex,
                      age,
                      ethnic_group,
                      IMD_quintile,
                      high_risk,
                      diseases_of_the_circulatory_system,
                      diseases_of_the_respiratory_system,
                      diseases_of_the_genitourinary_system,
                      diseases_of_the_digestive_system,
                      diseases_of_the_endocrine_system,
                      haematological_immunological_conditions,
                      infectious_diseases,
                      neurological_conditions,
                      cancers,
                      benign_neoplasm_cin,
                      mental_health_disorders,
                      musculoskeletal_conditions,
                      skin_conditions,
                      diseases_of_the_eye,
                      diseases_of_the_ear,
                      perinatal_conditions,
                      multimorbidity
                    FROM 
                      dars_nic_391419_j3w9t_collab.ccu013_paper_table_one_56million_denominator
                    WHERE
                      severity != 'no_covid'")

# EXCLUDE severity from query as this is a string therefore pushes memory size over 4.0 GB limit
# Separately extract males and females again due to OOM error
cohort_no_covid1 = dbGetQuery(con, 
                    "SELECT
                      death_covid,
                      death_all,
                      sex,
                      age,
                      ethnic_group,
                      IMD_quintile,
                      high_risk,
                      diseases_of_the_circulatory_system,
                      diseases_of_the_respiratory_system,
                      diseases_of_the_genitourinary_system,
                      diseases_of_the_digestive_system,
                      diseases_of_the_endocrine_system,
                      haematological_immunological_conditions,
                      infectious_diseases,
                      neurological_conditions,
                      cancers,
                      benign_neoplasm_cin,
                      mental_health_disorders,
                      musculoskeletal_conditions,
                      skin_conditions,
                      diseases_of_the_eye,
                      diseases_of_the_ear,
                      perinatal_conditions,
                      multimorbidity
                    FROM 
                      dars_nic_391419_j3w9t_collab.ccu013_paper_table_one_56million_denominator
                    WHERE
                      sex == 1
                    AND
                      severity == 'no_covid'")
# Note 3 distinct values for sex: 1, 2 & 'Unknown'
# Therefore use sex == 1 vs sex != 1 OR sex = 'Unknown'
cohort_no_covid2 = dbGetQuery(con, 
                    "SELECT
                      death_covid,
                      death_all,
                      sex,
                      age,
                      ethnic_group,
                      IMD_quintile,
                      high_risk,
                      diseases_of_the_circulatory_system,
                      diseases_of_the_respiratory_system,
                      diseases_of_the_genitourinary_system,
                      diseases_of_the_digestive_system,
                      diseases_of_the_endocrine_system,
                      haematological_immunological_conditions,
                      infectious_diseases,
                      neurological_conditions,
                      cancers,
                      benign_neoplasm_cin,
                      mental_health_disorders,
                      musculoskeletal_conditions,
                      skin_conditions,
                      diseases_of_the_eye,
                      diseases_of_the_ear,
                      perinatal_conditions,
                      multimorbidity
                    FROM 
                      dars_nic_391419_j3w9t_collab.ccu013_paper_table_one_56million_denominator
                    WHERE
                      (sex !=1 OR sex = 'Unknown')
                    AND
                      severity == 'no_covid'")
# Join sexes
cohort_no_covid = union_all(cohort_no_covid1,
                            cohort_no_covid2)
rm(cohort_no_covid1,
   cohort_no_covid2)
# Add severity
cohort_no_covid %<>% 
  mutate(severity = "no_covid")

# Create master 56 million cohort
cohort = union_all(cohort_covid, 
                   cohort_no_covid)

rm(cohort_no_covid,
   cohort_covid)
```


```{r process}
cohort %<>% 
  mutate(sex = case_when(sex == '1' ~ 'Male', 
                         sex == '2' ~ 'Female',
                         sex == 'Unknown' ~ 'Unknown')
         ) %>% 
  # Age: currently no NAs (encoded as 999) or Unknowns, else
  # mutate(age = na_if(age, 999)) %>% 
  mutate(age = as.numeric(age)) %>% 
  # Add binned ages, as per descriptive paper
  mutate(age_bin = case_when(age < 18 ~ "Under 18",
                           age <= 29 ~ "18-29",
                           age <= 49 ~ "30-49",
                           age <= 69 ~ "50-69",
                           age >= 70 ~ "70 or over")
       ) %>%  
  select(-age) %>% 
  # Factorise!
  mutate(across(c(sex, 
                  age_bin, 
                  ethnic_group, 
                  IMD_quintile, 
                  high_risk,
                  benign_neoplasm_cin,
                  cancers,
                  diseases_of_the_circulatory_system,
                  diseases_of_the_digestive_system,
                  diseases_of_the_ear,
                  diseases_of_the_endocrine_system,
                  diseases_of_the_eye,
                  diseases_of_the_genitourinary_system,
                  diseases_of_the_respiratory_system,
                  haematological_immunological_conditions,
                  infectious_diseases,
                  mental_health_disorders,
                  musculoskeletal_conditions,
                  neurological_conditions,
                  perinatal_conditions,
                  skin_conditions),
                as.factor)
         ) %>% 
  # re-level
  mutate(ethnic_group = relevel(ethnic_group, ref='White')
         ) %>% 
  # Factorise outcomes to get counts & % instead of mean(sd)
  mutate(across(c(severity,
                  death_covid,
                  death_all),
                as.factor)
         ) %>% 
  # Re-order to desired structure for tables
  relocate(death_covid,
           death_all,
           sex,
           age_bin,
           ethnic_group,
           IMD_quintile,
           high_risk,
           diseases_of_the_circulatory_system,
           diseases_of_the_respiratory_system,
           diseases_of_the_genitourinary_system,
           diseases_of_the_digestive_system,
           diseases_of_the_endocrine_system,
           haematological_immunological_conditions,
           infectious_diseases,
           neurological_conditions,
           cancers,
           benign_neoplasm_cin,
           mental_health_disorders,
           musculoskeletal_conditions,
           skin_conditions,
           diseases_of_the_eye,
           diseases_of_the_ear,
           perinatal_conditions,
           multimorbidity
           )
```




Check distribution of multimorbidity
```{r multimorbidity_dist}
hist(cohort$multimorbidity, breaks=30)
```

## All 56 million: COVID & No COVID
```{r population}
cohort %>%
  select(-severity) %>% 
  CreateTableOne(data=.) %>% 
  print(., 
        nonnormal = 'multimorbidity', 
        format.args = list(big.mark = ",")
        ) %>% 
  kableone(., 
           format.args = list(big.mark = ",")
           )
```

## Stratified by severity
```{r population_severity}
cohort %>%
  CreateTableOne(data=.,
                 strata = "severity") %>% 
  print(., 
        nonnormal = 'multimorbidity', 
        format.args = list(big.mark = ",")
        ) %>% 
  kableone(., 
           format.args = list(big.mark = ",")
           )
```

## COVID ONLY
```{r covid_only}
cohort %>%
  filter(severity != 'no_covid') %>% 
  select(-severity) %>% 
  CreateTableOne(data=.) %>% 
  print(., 
        nonnormal = 'multimorbidity', 
        format.args = list(big.mark = ",")
        ) %>% 
  kableone(., 
           format.args = list(big.mark = ",")
           )
```

## ALL COVID Deaths ONLY
As stratifying by severity only give those individuals where deaths was their only healthcare presentation, we want to show the demographics of all the COVID-19 deaths too.
```{r all_covid_deaths}
cohort %>%
  filter(death_covid == 1) %>% 
  select(-severity) %>% 
  CreateTableOne(data=.) %>% 
  print(., 
        nonnormal = 'multimorbidity', 
        format.args = list(big.mark = ",")
        ) %>% 
  kableone(., 
           format.args = list(big.mark = ",")
           )
```