---
title: 'Table 2: Characteristics of individuals experiencing COVID-19 events, stratified by wave'
author: 'Chris Tomlinson'
date: 'Knitted on `r date()`'

output:
  html_document:
    df_print: paged
---

**Description** 

This notebook aggregates the data within `ccu013_covid_events_demographics_paper_cohort` and stratifies per wave, to produce a series of 'Table 1's.  

  
**Project(s)** CCU013

**Paper** Characterising COVID-19 related events in a nationwide electronic health record cohort of 55.9 million people in England
 
**Author(s)** Chris Tomlinson
 
**Reviewer(s)** 
 
**Date last updated** 2021-09-29
 
**Date last reviewed** *NA*
 
**Date last run** `r date()`
 
**Data input**  
* `ccu013_covid_events_demographics_paper_cohort` 1 row per patient

**Data output**  
* Self-contained as the knitted `.html` file from this notebook

**Software and versions** `SQL`, `R`
 
**Packages and versions**

```{r config}
# Load libraries prior to sessionInfo() to ensure documented
library(dplyr)
library(tidyr)
library(magrittr)
library(stringr)
library(tableone)

warning("Requires authenticated connection to Spark saved as 'con'")
source("/mnt/efs/c.tomlinson/dbConnect.R")
```

```{r sessionInfo}
sessionInfo()
```

# 1. Load cohort & pre-process

For wave comparison we will retain `date_first` and split on this


```{r cohort}
cohort = dbGetQuery(con, 
                    "SELECT
                      date_first,
                      01_Covid_positive_test as positive_test,
                      01_GP_covid_diagnosis as primary_care_diagnosis, 
                      02_Covid_admission as hospitalisation, 
                      03_ECMO_treatment as ECMO, 
                      03_ICU_admission as ICU, 
                      03_IMV_treatment as IMV, 
                      03_NIV_treatment as NIV, 
                      04_Covid_inpatient_death as death_inpatient,
                      04_Fatal_with_covid_diagnosis as death_covid_dx,
                      04_Fatal_without_covid_diagnosis as death_no_covid_dx,
                      ventilatory_support,
                      death,
                      sex,
                      age,
                      ethnic_group,
                      IMD_quintile,
                      high_risk,
                      diseases_of_the_circulatory_system,
                      diseases_of_the_respiratory_system,
                      diseases_of_the_genitourinary_system,
                      diseases_of_the_digestive_system,
                      diseases_of_the_endocrine_system,
                      haematological_immunological_conditions,
                      infectious_diseases,
                      neurological_conditions,
                      cancers,
                      benign_neoplasm_cin,
                      mental_health_disorders,
                      musculoskeletal_conditions,
                      skin_conditions,
                      diseases_of_the_eye,
                      diseases_of_the_ear,
                      perinatal_conditions,
                      multimorbidity
                    FROM 
                      dars_nic_391419_j3w9t_collab.ccu013_covid_events_demographics_paper_cohort")


cohort %<>% 
  mutate(sex = case_when(sex == '1' ~ 'Male', 
                         sex == '2' ~ 'Female',
                         sex == 'Unknown' ~ 'Unknown')
         ) %>%  
  # Age 
  # Currently no NAs (encoded as 999) or Unknowns 
  # mutate(age = na_if(age, 999)) %>% 
  mutate(age = as.numeric(age)) %>% 
  # Add binned ages, as per descriptive paper
  mutate(age_bin = case_when(age < 18 ~ "Under 18",
                           age <= 29 ~ "18-29",
                           age <= 49 ~ "30-49",
                           age <= 69 ~ "50-69",
                           age >= 70 ~ "70 or over")
       ) %>%  
  select(-age) %>% 
  # Factorise!
  mutate(across(c(sex, 
                  age_bin, 
                  ethnic_group, 
                  IMD_quintile, 
                  high_risk, 
                  benign_neoplasm_cin,
                 cancers,
                 diseases_of_the_circulatory_system,
                 diseases_of_the_digestive_system,
                 diseases_of_the_ear,
                 diseases_of_the_endocrine_system,
                 diseases_of_the_eye,
                 diseases_of_the_genitourinary_system,
                 diseases_of_the_respiratory_system,
                 haematological_immunological_conditions,
                 infectious_diseases,
                 mental_health_disorders,
                 musculoskeletal_conditions,
                 neurological_conditions,
                 perinatal_conditions,
                 skin_conditions),
                as.factor)
         ) %>% 
  # re-level
  mutate(ethnic_group = relevel(ethnic_group, ref='White')
         ) %>% 
  # OUTCOMES
    # Factorise outcomes to get counts & % instead of mean(sd)
  mutate(across(c(death,
                  ventilatory_support,
                  positive_test, 
                  primary_care_diagnosis, 
                  hospitalisation, 
                  ICU,
                  NIV,
                  IMV,
                  ECMO,
                  death_inpatient,
                  death_covid_dx,
                  death_no_covid_dx
                  ),
                as.factor)
         ) %>% 
  # re-order columns to get desired layout of table 1 rows
  relocate(# Outcomes first
          positive_test, 
                  primary_care_diagnosis, 
                  hospitalisation, 
                  ventilatory_support,
                  ICU,
                  NIV,
                  IMV,
                  ECMO,
                  death,
                  death_inpatient,
                  death_covid_dx,
                  death_no_covid_dx,
          # Demographics
          sex,
          age_bin,
          ethnic_group,
          IMD_quintile,
          # Comorbidities
          high_risk,
          diseases_of_the_circulatory_system,
          diseases_of_the_respiratory_system,
          diseases_of_the_genitourinary_system,
          diseases_of_the_digestive_system,
          diseases_of_the_endocrine_system,
          haematological_immunological_conditions,
          infectious_diseases,
          neurological_conditions,
          cancers,
          benign_neoplasm_cin,
          mental_health_disorders,
          musculoskeletal_conditions,
          skin_conditions,
          diseases_of_the_eye,
          diseases_of_the_ear,
          perinatal_conditions,
          multimorbidity,
          # Outcomes
          death
          )
```

```{r multimorbidity_dist}
hist(cohort$multimorbidity)
```


# All Cases (no waves)
```{r}
cohort %>%
  select(-date_first) %>% 
  CreateTableOne(data=.) %>% 
  print(., 
        nonnormal = 'multimorbidity', 
        format.args = list(big.mark = ",")
        ) %>% 
  kableone(., 
           format.args = list(big.mark = ",")
           )
```

## ALL COVID-19 Hospitalisations
```{r}
cohort %>%
  filter(hospitalisation == 1) %>% 
  CreateTableOne(data=.) %>% 
  print(., 
        nonnormal = 'multimorbidity', 
        format.args = list(big.mark = ",")
        ) %>% 
  kableone(., 
           format.args = list(big.mark = ",")
           )
```

### Hospitalisations without ventilatory_support

```{r}
cohort %>%
  filter(hospitalisation == 1,
         ventilatory_support == 0) %>% 
  CreateTableOne(data=.) %>% 
  print(., 
        nonnormal = 'multimorbidity', 
        format.args = list(big.mark = ",")
        ) %>% 
  kableone(., 
           format.args = list(big.mark = ",")
           )
```
## ALL ICU admissions
```{r}
cohort %>%
  filter(ICU == 1) %>% 
  CreateTableOne(data=.) %>% 
  print(., 
        nonnormal = 'multimorbidity', 
        format.args = list(big.mark = ",")
        ) %>% 
  kableone(., 
           format.args = list(big.mark = ",")
           )
```

## ALL ventilatory support
```{r}
cohort %>%
  filter(ventilatory_support == 1) %>% 
  CreateTableOne(data=.) %>% 
  print(., 
        nonnormal = 'multimorbidity', 
        format.args = list(big.mark = ",")
        ) %>% 
  kableone(., 
           format.args = list(big.mark = ",")
           )
```

## ALL ventilatory support OUTSIDE ICU
```{r}
cohort %>%
  filter(ICU == 0,
         ventilatory_support == 1) %>% 
  CreateTableOne(data=.) %>% 
  print(., 
        nonnormal = 'multimorbidity', 
        format.args = list(big.mark = ",")
        ) %>% 
  kableone(., 
           format.args = list(big.mark = ",")
           )
```

## ALL COVID-19 Deaths (no waves)
```{r}
cohort %>%
  filter(death == 1) %>% 
  CreateTableOne(data=.) %>% 
  print(., 
        nonnormal = 'multimorbidity', 
        format.args = list(big.mark = ",")
        ) %>% 
  kableone(., 
           format.args = list(big.mark = ",")
           )
```


# Split into waves

```{r wave_split}
wave_1_start = "2020-03-20"
wave_1_end = "2020-05-29"
wave_2_start = "2020-09-30"
wave_2_end = "2021-02-12"

cohort %<>% 
  mutate(wave = if_else(date_first >= wave_1_start & 
                          date_first <= wave_1_end, 
                        1, 1.5) # name inter-wave period as 1.5
         ) %>%
  mutate(wave = if_else(date_first >= wave_2_start & 
                          date_first <= wave_2_end, 
                        2, wave)
         ) %>% 
  select(-date_first)
```

# All patients, stratified by wave  
* NB vastly different numbers: uneven durations + much greater testing  
* Therefore this is not used in the manuscript at present but included here for completion  
```{r}
cohort %>%
  filter(wave != 1.5) %>% 
  CreateTableOne(data=.,
                 strata = "wave") %>% 
  print(., 
        nonnormal = 'multimorbidity', 
        format.args = list(big.mark = ",")
        ) %>% 
  kableone(., 
           format.args = list(big.mark = ",")
           )
```

# Subset to ALL hospitalised patients, stratified by wave
This should be less sensitive to the increase in testing  
```{r}
cohort %>%
  filter(wave != 1.5,
         hospitalisation == 1) %>% 
  CreateTableOne(data=.,
                 strata = "wave") %>% 
  print(., 
        nonnormal = 'multimorbidity', 
        format.args = list(big.mark = ",")
        ) %>% 
  kableone(., 
           format.args = list(big.mark = ",")
           )
```

## Subset to hospitalised patients who *don't* receive any ventilatory support
```{r}
cohort %>%
  filter(wave != 1.5,
         hospitalisation == 1,
         ventilatory_support == 0) %>% 
  CreateTableOne(data=.,
                 strata = "wave") %>% 
  print(., 
        nonnormal = 'multimorbidity', 
        format.args = list(big.mark = ",")
        ) %>% 
  kableone(., 
           format.args = list(big.mark = ",")
           )
```


# Subset to only `ventilatory_support` patients, stratified by wave  
```{r}
cohort %>%
  filter(wave != 1.5,
         ventilatory_support == 1) %>% 
  CreateTableOne(data=.,
                 strata = "wave") %>% 
  print(., 
        nonnormal = 'multimorbidity', 
        format.args = list(big.mark = ",")
        ) %>% 
  kableone(., 
           format.args = list(big.mark = ",")
           )
```

## ICU admissions only
```{r}
cohort %>%
  filter(wave != 1.5,
         ICU == 1) %>% 
  CreateTableOne(data=.,
                 strata = "wave") %>% 
  print(., 
        nonnormal = 'multimorbidity', 
        format.args = list(big.mark = ",")
        ) %>% 
  kableone(., 
           format.args = list(big.mark = ",")
           )
```


## ventilatory support OUTSIDE ICU only
```{r}
cohort %>%
  filter(wave != 1.5,
         ventilatory_support == 1,
         ICU == 0) %>% 
  CreateTableOne(data=.,
                 strata = "wave") %>% 
  print(., 
        nonnormal = 'multimorbidity', 
        format.args = list(big.mark = ",")
        ) %>% 
  kableone(., 
           format.args = list(big.mark = ",")
           )
```

#  Subset to only deceased patients, stratified by wave  
```{r}
cohort %>%
  filter(wave != 1.5,
         death == 1) %>% 
  CreateTableOne(data=.,
                 strata = "wave") %>% 
  print(., 
        nonnormal = 'multimorbidity', 
        format.args = list(big.mark = ",")
        ) %>% 
  kableone(., 
           format.args = list(big.mark = ",")
           )
```


# Additional workflow steps  
To my knowledge, `tableOne()` doesn't allow for multiple groups & stratification. Workaround is therefore to copy and paste the output above into google sheets next to each other and then in to the manuscript. 